---
title: "ch1_climate_change"
output:  
  md_document:
    toc: true
    variant: markdown_github
bibliography: ../refs/references.bib
csl: ../refs/ecology.csl
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup, echo=FALSE}
show_text <- FALSE
```


# Clima en SN 


Notas para cambio en el clima en montañas: 
* Ver datos de Nogués-Bravo et al. 2008 


Mas notas sobre sequía en ambientes pasados (holoceno y demás) 
The progressive middle and late Holocene trend toward arid conditions identified by numerous authors in the western Mediterranean region, mostly related to a decrease in summer insolation, is also documented in this record



Our multiproxy record shows little evidence of human impact in the area until ∼ 1550 cal yr BP, when evidence of agriculture and livestock grazing occurs. Therefore, climate is the main forcing mechanism controlling environmental change in the area until relatively recently.

Ramos-Román et al. 2018 


Our work confirms the progressive aridification trend that occurred
during at least the last ∼ 4700 cal yr BP on the southern
Iberian Peninsula, as shown here by the progressive decrease
in Mediterranean forest and the increase in herbs


This natural
progressive aridification confirmed by the decrease in
Mediterranean forest taxa


A clear increase in human land use is also observed during the last ∼ 1550 cal yr BP (see bellow), including abundant Glomus from erosion, which shows that humans were at least partially responsible for this sedimentary change

A suite of proxy previous studies supports our conclusions
regarding the aridification trend since the middle Holocene
(Carrión, 2002; Carrión et al., 2010; Fletcher and SánchezGoñi,
2008; Fletcher et al., 2013; Jiménez-Espejo et al.,
2014; Jiménez-Moreno et al., 2015)


 In the western Mediterranean
region the decline in forest development during the
middle and late Holocene is related to a decrease in summer
insolation (Jiménez-Moreno and Anderson, 2012; Fletcher
et al., 2013), 


The
aridification trend can clearly be seen in the nearby alpine
records from the Sierra Nevada, where there was little influence
from human activity (Anderson et al., 2011; JiménezMoreno
and Anderson, 2012; Jiménez-Moreno et al., 2013;
Ramos-Román et al., 2016)


Humans have probably had an impact on the area since prehistoric
times; however, the Padul-15-05 multiproxy record
shows a more significant human impact during the last
∼ 1550 cal yr BP, which intensified in the last ∼ 500 years
(since 1450 CE to present). This is deduced by a significant
increase in nitrophilous plant taxa such as Cichorioideae,
Convolvulaceae, Polygonaceae, and Plantago and
the increase in some NPPs such as Tilletia, coprophilous
fungi, and thecamoebians 



Nevertheless, a study in the southeastern part of the
Iberian Peninsula (Bellin et al., 2013) suggested that major
soil erosion could have occurred from the abandonment of
agricultural activities in the mountain areas as well as the
abandonment of irrigated terrace systems during the Christian
Reconquest. Enhanced soil erosion at this time is also
supported by the increase in Glomus type (Figs. 4 and 8).

------ 
# Notas sobre sequía. `TODO` Ordenar 


definir de forma genérica las sequías
Las sequías son un fenómeno ... 

Following Mishra and Singh (2010) drought can be classified as meteorological, agricultural, hydrological, ground-water and socio-economic. ---> Mirar 

... hablar de que es algo típico en el mediterráneo, pero que en los últimos años está ocurriendo un aumento debido al cambio climático ... 

... aquí revisamos un poco el estado del arte (estudios )

Varios estudios, utilizando diferentes fuentes de datos e índices de sequía, han puesto de manifiesto la existencia de un incremento en la frecuencia de las sequías [@Lloyd2002; @Sousa2011; @VicenteSerrano2014; @Coll2016]. Durante el siglo XX se ha observado una tendencia hacia condiciones mas sécas en Europa [@Linolello2012; @Spinoni2015], siendo el incremento de la frecuencia de las sequías particularmente importante para la región Mediterránea [@Hoerling2012]. En la Penísula Ibérica, además se ha observado un aumento en la severidad de las sequías [@VicenteSerrano2014; @Pascoa2017], especialmente en la zona sur donde 

These agree with the findings of Greve et al. [38] that identified the Southwest of the IP as one hot spot of the pattern “dry gets drier.”
conviertiendo a la zona sur de Europa en un *hotspot* del cambio climático [@Greve2014]. 


Severe and long drought events are frequent in the IP, with strong effects in ecosystems and the related social and economic impacts [10, 18]. 




... y luego aquí hablamos de Sierra Nevada en cuestión (con datos que analicemos) 

# CRU 
El conjunto de datos CRU procede de cientos de estaciones dispersas de forma no aleatoria a lo largo de la tierra, con mayor cantidad de estaciones en latitudes medias que en las zonas tropicales o en el Ártico. 

Algunos autores [Harris et al] han comprobado una alta correlación de las series de datos CRU para precipitación y temperatura con otros conjuntos de datos climáticos. 

Este conjunto de datos con resolución 0.5 x 0.5 º se ha recomendado que es útil para diferentes aplicaciones climáticas: modelado hidrológico, escenarios de cambio climático y evaluación de modelos regionales [@Mitchell2005]. 




En la PI se han detectado los siguientes patrones: 
1940's, 1950's y desde 1980's los eventos de sequía mas largos y severos. 
1960's y 1970's fueron predominantemente húmedos 

Además en estas decadas la mayor parte del área de la PI estaba con valores de sequía moderados (spei < -)









Se ha identificado una clara tendencia hacia condiciones mas sécas durante el siglo XX en la región Mediterránea de Europa, incluyendo la p

En la Peninsula Iberica durante el siglo XX se ha observado una tendencia hacia condiciones mas secas [@
Pascoa2017; 

]





En el sur de Europa, en particular en la región Mediterránea, se ha observado un incremento en la frecuencia de las sequías [@Hoerling2012]. 

Varios trabajos han puesto de manifiesto el incremento en la frecuencia de las sequías para la región Mediterránea []
--- 






[@VicenteSerrano2014]
S. M. Vicente-Serrano, J.-I. Lopez-Moreno, S. Beguería et al., “Evidence of increasing drought severity caused by temperature rise in southern Europe,” Environmental Research Letters, vol. 9, no. 4, Article ID 044001, 2014.










@Pascoa2017: 
https://www.hindawi.com/journals/amete/2017/4653126/#B24

@Lionello2012
P. Lionello, Ed., The Climate of the Mediterranean Region: From the Past to the Future, Elsevier, 2012.

@Spinoni2015
J. Spinoni, G. Naumann, J. Vogt, and P. Barbosa, “European drought climatologies and trends based on a multi-indicator approach,” Global and Planetary Change, vol. 127, pp. 50–57, 2015.

@Coll2016
J. Coll, E. Aguilar, and L. Ashcroft, “Drought variability and change across the Iberian Peninsula,” Theoretical and Applied Climatology, vol. 130, no. 3-4, pp. 901–916, 2016. 

@Hoerling2012
Hoerling,M., Eischeid, J., Perlwitz, J., Quan, X., Zhang, T., Pegion, P., 2012. On the increased
frequency of Mediterranean drought. J. Clim. 25 (6), 2146–2161.

@Lloyd2002
B. Lloyd-Hughes and M. A. Saunders, “A drought climatology for Europe,” International Journal of Climatology, vol. 22, no. 13, pp. 1571–1592, 2002

@Sousa2011 
P. M. Sousa, R. M. Trigo, P. Aizpurua, R. Nieto, L. Gimeno, and R. Garcia-Herrera, “Trends and extremes of drought indices throughout the 20th century in the Mediterranean, Natural Hazards and Earth System Sciences, vol. 11, no. 1, pp. 33–51, 2011

@Greve2014
P. Greve, B. Orlowsky, B. Mueller, J. Sheffield, M. Reichstein, and S. I. Seneviratne, “Global assessment of trends in wetting and drying over land,” Nature Geoscience, vol. 7, no. 10, pp. 716–721, 2014

@Mitchell2005
T. D. Mitchell and P. D. Jones, “An improved method of constructing a database of monthly climate observations and associated high-resolution grids,” International Journal of Climatology, vol. 25, no. 6, pp. 693–712, 2005


```{r echo=FALSE, results='asis', eval=show_text}
cat(" 

")
```




